---
title: "Assignment 2: Multinomial Choices"
author: "Yongyi Lin"
date: "3/24/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
library(tidyverse)
library(bayesm)
```

```{r dataset}
setwd("/Users/yongyilin/Econ613/Assignments/A2")
data(margarine)
```

## Exercise 1 Data Description
```{r Exercise 1}
choiceprice <- margarine$choicePrice
demos <- margarine$demos
### Average and dispersion in product characteristics
prod_avg <- as.matrix(apply(choiceprice[,2:12], 2, mean))
prod_avg
prod_disp <- as.matrix(apply(choiceprice[,2:12], 2, sd))
prod_disp
### Market share and market share by product characteristics
prod_share = as.matrix(summary(as.factor(choiceprice[,2])))/nrow(choiceprice)
prod_share

mar <- merge(x = choiceprice, y = demos, by = "hhid", all.x = TRUE)
### Mapping between income and choice
table(mar[,c(2,13)])
### Mapping between family size and choice
table(mar[,c(2,14)])
table(mar[,c(2,15)])
table(mar[,c(2,16)])
### Mapping between education status and choice
table(mar[,c(2,17)])
### Mapping between job status and choice
table(mar[,c(2,18)])
### Mapping between retirement status and choice
table(mar[,c(2,19)])
```


```{r}
choice <- 1:10
names(choice) <- 1:10
y <- as.matrix(map_df(choice, function(x) as.integer(choiceprice$choice == x)))
```
## Exercise 2 First Model
### Conditional Logit Model
```{r Exercise 2}
# Exercise 2 First Model
# Conditional logit model
x_1 <- mar[,3:12]
conl_p <- function(x,b) {
  pn <- exp(matrix(rep(c(0,b[1:9]),nrow(x)),byrow = TRUE,nrow(x))+x*b[10])
  pd <- apply(pn,1,sum)
  return(pn/pd)
}
conl_ll <- function(y,x,b) {
  l <- -sum(y*log(conl_p(x,b)))
  return(l)
}
conl <- optim(function(b) conl_ll(y=y,x=x_1,b=b),par=rep(0,10),method="BFGS")
conl$par
```
#### The 10th parameter refers to the effect of price. The negative coefficient (-6.6566340) indicates that individual would be less likely to purchase the product as the price increases.

## Exercise 3 Second Model
### Multinomial Logit Model
```{r Exercise 3}
# Exercise 3 Second Model
# Multinomial logit model
x_2 <- as.matrix(mar[,13],ncol=1)
colnames(x_2)[1] <- "income"
multil_p <- function(x,b) {
  pn <- exp(matrix(rep(c(0,b[1:9]),nrow(x)),byrow=TRUE,nrow(x))
                   +t(apply(x,1,function(x)x*c(0,b[10:18]))))
  pd <- matrix(apply(pn,1,sum)) %*% t(rep(1,10))
  return(pn/pd)
}
multil_ll <- function(y,x,b) {
  l <- -sum(y*log(multil_p(x,b)))
  return(l)
}
multil <- optim(function(b) multil_ll(y=y,x=x_2,b=b),par=rep(0,18),method="BFGS")
multil$par
```
#### The last 9 parameters are the income effects of products 2~10. Thus, an individual would be more likely to purchase product 3, 4, 6, 8, 9, and 10, and less likely to purchase product 2, 5, and 7.

## Exercise 4 Marginal Effects
### Conditional Logit Model
```{r Exercise 4_1}
# Marginal effect for the first model
p_conl <- conl_p(x_1,conl$par)
ind <- array(0, dim=c(nrow(x_1),ncol(x_1),ncol(x_1)))
for (i in 1:nrow(x_1)) {
  diag(ind[i,,]) <- 1
}
me_conl <- array(0, dim=c(nrow(x_1),ncol(x_1),ncol(x_1)))
for (i in 1:nrow(x_1)) {
  for (j in 1:ncol(x_1)) {
    for (k in 1:ncol(x_1)) {
      me_conl[i,j,k] <- p_conl[i,j]*(ind[i,j,k]-p_conl[i,k])*conl$par[10]
    }
  }
}
apply(me_conl,c(2,3),mean)
```
#### As one can see from the matrix, only diagonal elements are negative whereas all the other elements are positive. This indicates that an individual would turn to other products, if the price of a given product increased, which is very intuitive and makes perfect sense.

### Multinomial Logit Model
```{r Exercise 4_2}
# Marginal effect for the second model
p_multil <- multil_p(x_2,multil$par)
b_multil <- c(0,multil$par[10:18])
me_multil <- array(0,dim=c(nrow(x_2),10))
for (i in 1:nrow(x_2)) {
  b_bar <- sum(p_multil[i,]*b_multil)
  for (j in 1:10) {
    me_multil[i,j] <- p_multil[i,j]*(b_multil[j]-b_bar)
  }
}
for (i in 1:nrow(x_2)) {
  b_bar <- sum(p_multil[i,]*b_multil)
  me_multil[i,] <- p_multil[i,]*(b_multil-b_bar)
}
apply(me_multil,2,mean)
```
#### In this model, an individual would purchase product 1, 2, 5, and 7 more if his or her income increased.

## Exercise 5 IIA
```{r Exercise 5}
mix_ll <- function(y,x,b,p_mix) {
  return(-sum(y*log(p_mix(x,b))))
}
# Mixed logit with all choices
X_1 <- as.matrix(mar[,3:13])
p_mix_1 <- function(x,b) {
  pn <- exp(
    matrix(rep(c(0,b[1:9]),nrow(x)),
           byrow=TRUE,
           nrow(x))
    +x[,1:10]*b[10]
    +t(apply(matrix(x[,11],ncol=1),1,function(x) x*c(0,b[11:19])))
  )
  pd <- apply(pn,1,sum)
  return(pn/pd)
}
mixl_1 <- optim(function(b) mix_ll(y=y,x=X_1,b=b,p_mix=p_mix_1),par=rep(0,19),method="BFGS")
mixl_1$par
# Alternative specification: remove the fifth choice
X_2 <- X_1[,-5]
p_mix_2 <- function(x,b) {
  pn <- exp(
    matrix(rep(c(0,b[1:8]),nrow(x)),byrow=TRUE,nrow(x))
    +x[,1:9]*b[9]
    +t(apply(matrix(x[,10],ncol=1),1,function(x) x*c(0,b[10:17])))
  )
  pd <- apply(pn,1,sum)
  return(pn/pd)
}
mixl_2 <- optim(function(b) mix_ll(y=y[,-5],x=X_2,b=b,p_mix=p_mix_2),par=rep(0,17),method="BFGS")
mixl_2$par
# Compute test statistics
L_all <- mix_ll(y=y,x=X_1,b=mixl_1$par,p_mix=p_mix_1)
L_alter <- mix_ll(y=y[,-2],x=X_2,b=mixl_2$par,p_mix=p_mix_2)
MTT <- 2*(L_all-L_alter)
csq95 <- qchisq(.95, length(mixl_2$par))
MTT > csq95
```
#### MTT > csq95. IIA is violated.